# Exercise 1: Data Visualisation

## Part 1

### 1.1 Explore the mpg dataset



Start a new R-script within your R-project, and load the library 'tidyverse'. 

```{r, message = FALSE, warning = FALSE, }
# load packages
library(tidyverse)
```

Look at the data:

```{r, eval = FALSE }
mpg
```

Explore the mpg dataset, using functions like `summary()`, `dim()`, `head()` to answer the following questions:

1. How many rows and columns does it have?
2. What do the different variables describe? Inspect using the R-help: `?mpg`
3. What is the difference between  variables like e.g. 'model', 'trans' and 'displ', 'hwy'?


### 1.2 Create ggplot-figures

1. Create a scatterplot of 'hwy' vs. 'cyl'. 
2. Save the plot as pdf and find it in your working directory
3. Make a plot of 'manufacturer' vs. 'class'. Why is this plot not useful? Could you think of more useful ways to display such a relation?



### 1.3. Investigate 'Outliers'

Fuel efficiency (hwy) declines with enginge size (displ). However, there is a few cars that have a higher efficiency than expected from this general trend (highlighted in red). Can you explain these cars?

Use your knowledge of the mpg-dataset and your ggplot skills to identify this group of cars.

```{r, echo = FALSE, out.width = "60%"}
mpg %>%
  mutate(id = ifelse(class == "2seater", "A", "B")) %>%
  ggplot(aes(x = displ, y = hwy, color = id)) +
  geom_point() +
  scale_color_manual(values = c("red", "black")) +
  theme(legend.position = "none")
```


## Part 2

### 2.1 Using aestethics: `aes()`

Recreate the scatterplot of displ vs. hwy. Include further variables making use of additional mappings, e.g. color, size or shape. 

* Try to include information on the number of cyclinder (cyl), the type of car (class). What pattern can you identify?


```{r, include = FALSE}

mpg %>%
  ggplot(aes(x = displ, y = hwy, color = class)) +
  geom_point(aes(size = cyl))
```


* Can you use the different aesthetics for all types of variables? 
What type of aesthetics are useful for which type of variables? 

```{r, include = FALSE, eval = FALSE}

mpg %>%
  ggplot(aes(x = displ, y = hwy, color = class)) +
  geom_point(aes(size = cyl))

## produces error, because shape needs to be discrete
mpg %>%
  ggplot(aes(x = displ, y = hwy, color = class)) +
  geom_point(aes(shape = cyl))

mpg %>%
  ggplot(aes(x = displ, y = hwy, color = class)) +
  geom_point(aes(shape = drv))
```


* Create the two figures below and check out the color scales. What is the difference and what is the reason for it? What happens if you substitute `color = year` by `color = as.factor(year)`?

```{r, eval = FALSE}
mpg %>%
  ggplot(aes(x = displ, y = hwy, color = year)) +
  geom_point()

mpg %>%
  ggplot(aes(x = displ, y = hwy, color = drv)) +
  geom_point()
```


### 2.2 Moving on beyond the scatter


* Summarize the number of cars of each manufacturer in a figure and try to take the different classes into account. 
Hint: `geom_bar()` might be your friend.

```{r, include = FALSE}

ggplot(data = mpg, aes(manufacturer, fill = class)) +
  geom_bar()
```


* Boxplots are really useful since they contain a lot of information. Use boxplots to investigate whether different different fuel types ('fl') have different efficiency (hwy). 

* Use facets to go into more detail. 

```{r, include = FALSE}
mpg %>%
  ggplot(aes(x = fl, y = hwy)) +
  geom_boxplot()

mpg %>%
  ggplot(aes(x = fl, y = hwy)) +
  geom_boxplot() +
  facet_wrap(~class)
```






