# Exercise: Tidy Data

## Part 1: Which data structure is tidy?

Check out these datasets and try to answer the following questions:

```{r, eval = FALSE}
library(tidyverse)
table1
table2
table3
table4a
table4b
```


- how are they organised? 
- which fulfills the criteria of tidy data? why? or why not?
- you want to calculate the infection rate = cases / population * 1000. Describe in prose how you would do it based on the different tables?

- __Which is easiest to work with? __


## Part 2: Transform tables 

Reproduce the following figure starting with `table1`,` table2`, and `table3`. 

Calculate rate = cases/population * 1000

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
table1 %>% 
  mutate(rate = cases/population * 1000) %>% 
  pivot_longer(cols = c(cases:rate), 
               names_to = "Variable", 
               values_to = "Value") %>% 
  ggplot(aes(x = year, y = Value, color = country)) + 
  geom_point() + geom_line() +
  facet_grid(Variable ~., scales = "free")
```


:::{.callout-tip collapse="true"}

## Solution

```{r, eval = FALSE, include = TRUE}
table1 %>% 
  mutate(rate = cases/population * 1000) %>% 
  pivot_longer(cols = c(cases:rate), 
               names_to = "Variable", 
               values_to = "Value") %>% 
  ggplot(aes(x = year, y = Value, color = country)) + 
  geom_point() + 
  facet_wrap(Variable ~., scales = "free")

table2 %>% 
  pivot_wider(names_from = "type", values_from = "count") %>% 
  mutate(rate = cases/population * 1000) %>% 
  pivot_longer(cols = c(cases:rate), 
               names_to = "Variable", 
               values_to = "Value") %>% 
  ggplot(aes(x = year, y = Value, color = country)) + 
  geom_point() + 
  facet_wrap(Variable ~., scales = "free")

table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE) %>% 
  mutate(rate = cases/population * 1000) %>% 
  pivot_longer(cols = c(cases:rate), 
               names_to = "Variable", 
               values_to = "Value") %>% 
  ggplot(aes(x = year, y = Value, color = country)) + 
  geom_point() + 
  facet_wrap(Variable ~., scales = "free")
  
  
```

:::
